Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-10 3.Thesis/Writeup/2.Manuscript.tex
Until p. 9

These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}. There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{taka17}. Moreover, the behavoir of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}. Therefore, algorithmic convergence should be monitored carefully, \textbf{but this is not straightforward}. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence \citep{gelm13}. Since MCMC algorithms do not reach a unique point at which convergence is established, \textbf{methods} may only identify signs of \emph{non}-convergence \citep{hoff09}. Several non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.

In this paper, we study different methods for assessing non-convergence in iterative imputation algorithms. We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications. We then adress the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in \texttt{R} \citep{R}. The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged. And additionally, to formulate an informed advice on when to safely conclude approximate convergence. We will therefore assess the performance of different methods to identify non-convergence. We translate the results of the study into guidelines for applied researchers. \textbf{ADD that these guidelines may aid to draw valid inference from incomplete data?{]}} For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.

@@

Let \(y\) denote an \(n \times k\) matrix containing the data values on \(k\) variables for all \(n\) units in a sample. The data value of unit \(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\) (\(j = 1, 2, \dots, k\)) may be either observed or missing. The number of units \(i\) with at least one missing value, divided by the total number of units \(n\), is called the missingness proportion \(p_{\rm mis}\) in dataset \(y\). The collection of observed data values in \(y\) is denoted by \(y_{\rm obs}\); the missing part of \(y\) is referred to as \(y_{\rm mis}\). For each datapoint in \(y_{\rm mis}\), we sample \(m \times T\) plausible values, where \(m\) is the number of imputations (\(\ell = 1, 2, \dots, m\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm. The state space of the algorithm at a certain iteration \(t\) may be summarized by a scalar summary \(\theta\) (e.g., \textbf{the average of the imputed values}). The collection of \(\theta\)-values between the \textbf{initial value} (at \(t=1\)) and the final imputed value (at \(t=T\)) will be referred to as an `imputation chain'.

@@

There are two requirements for convergence of iterative algorithms: mixing and stationarity \citep{gelm13}. In iterative imputation algorithms, mixing implies that imputation chains intermingle nicely, and stationarity is characterized by \textbf{the absence of trending} within chains. If one of these two requirements is not met, we speak of non-convergence. Without mixing, chains may be `stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values. The distribution of imputed values then differs across imputations. This may cause under-estimation of the variance between chains, which results in spurious, invalid inferences. Without stationarity, there is trending within imputation chains. Trending implies that further iterations would yield a systematically lower or higher set of imputations. Iterative imputation algorithms that have not (yet) reached stationarity, may thus yield biased estimates.

\hypertarget{non-convergence-in-mice}{%
\subsection{Non-convergence in MICE}\label{non-convergence-in-mice}}

To illustrate what non-mixing and non-stationarity looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure \ref{fig:non-conv} displays two scenarios from the example. The graph on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. Each line portrays one imputation chain: scalar summary \(\theta\) plotted against the iteration number \(t\). The \(\theta\) in this example is the average of the imputed values for variable \(j\) in each imputation \(\dot y_{\rm imp, \ell}\) (also called the `chain mean' of \(j\)). In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence scenario, there is a lot of trending and some chains do not intermingle. Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two scenarios. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. It is obvious that non-convergence impacts the distribution of the imputed values in this example. This effect may translate into the distribution of the completed data, which consequently affects the estimated quantity of scientific interest \(\bar{Q}\). \(\bar{Q}\) may be biased due to systematically higher or lower imputed values, and invalid because the variance within imputation chains is under-estimated. Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms. \textbf{{[}ADD t and l to each theta??{]}}

@@

\caption{Typical convergence versus pathological non-convergence. Please note that this is the same data with a different imputation model, leading to different imputations (e.g., see the units on the y-axis).}\label{fig:non-conv}
\end{figure}

@@

Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection. After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots. In a traceplot, a scalar summary of the state space of the algorithm \(\theta\) is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}. Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends \citep{buur18}. In \texttt{mice}, the default scalar summaries to inspect are chain means and chain variances.

Identifying non-convergence by inspecting traceplots may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence \citep[\(\S\) 6.5.2]{buur18}. Moreover, traceplots typically only display chain means and chain variances, which are univariate scalar summaries of the state space of the algorithm. Monitoring univariate \(\theta\)s may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of \(\dot y_{\rm imp, \ell}\). Ideally, a multivariate \(\theta\) should be monitored.

A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient), which usually is multivariate in nature. Implementing this, however, might be somewhat too technical for empirical researchers. Moreover, this scalar summary is not model-independent, i.e., it is not universal to all complete data problems. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest. Additionally, one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently. On these grounds, such a \(\theta\) can be considered insufficient too.

@@

The downside to the current \(\theta\)s to evaluate is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent \(\theta\) that summarizes the multivariate nature of the data.

We propose \(\lambda_{1, \ell}\) as such a scalar summary. We define \(\lambda_{1, \ell}\) as the first eigenvalue of the variance-covariance matrix of the completed data. Let \(\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}\) be the eigenvalues of \(\Sigma_\ell\) in each of the \(m\) completed datasets \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}. \(\lambda_{1, \ell}\) is measure that summarizes the covariances in the completed datasets. The first eigenvalue has the appealing property that is not dependent on the substantive model of interest. Eigenvalue decomposition is inspired by \citet{mack03}.\textbf{{[}ADD definition variance in first PCA component is equal to the first eigenvalue of the variance-covariance matrix{]}}

@@

Non-convergence diagnostics for MCMC algorithms typically identify problems in either one of the two requirements for convergence: mixing or stationarity. Non-stationarity within chains may be diagnosed with e.g., autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's \citeyearpar{raft91} procedure to determine the effect of trending on the precision of estimates. A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor \(\widehat{R}\) \citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed adaptation, \(\widehat{R}\) might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated \citep{veht19}. Therefore, we use \(\widehat{R}\) \textbf{{[}WHICH? Explain that we'll use both{]}} and \(AC\) to evaluate mixing and stationarity separately, as recommended by e.g., \citet{cowl96}.

\textbf{{[}Add: we don't use effective sample size and MC error because they assume independent samples \emph{within} chains. In MI, we only use the final estimate. Also, computing n\_eff assumes infinitely many iterations, which is fine in Bayesian analyses where T is often at least one thousand. But the default in iterative imp is often substantially lower.{]}}

@@

The potential scale reduction factor \(\widehat{R}\) was coined by \citet{gelm92}. An updated version of \(\widehat{R}\) has been proposed by Vehtari et al.~\citeyearpar[p.~5]{veht19}. \textbf{This version is better suited to detect non-convergence in the tails of distributions, by using three transformations on the scalar summary that it is applied to. Namely, rank-normalization, folding, and localization. The Vehtari et al.~version of \(\widehat{R}\) may be suitable for iterative imputation. We therefore use their definition of the diagnostic.} Let \(m\) be the total number of chains, \(T\) the number of iterations per chain (where \(T\geq2\)), and \(\theta\) the scalar summary of interest. For each chain (\(\ell = 1, 2, \dots, m\)), we estimate the variance of \(\theta\), and average these to obtain within-chain variance \(W\).

@@

\noindent We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\)s per chain. \textbf{{[}Note that this is not the usual B in MI{]}}

@@

Autocorrelation is defined as the correlation between two subsequent \(\theta\)-values within the same chain \citep[p.~147]{lync07}. In this study, we only consider \(AC\) at lag 1, i.e., the correlation between the \(t^{th}\) and \(t+1^{th}\) iteration of the same chain. Following the same notation as for \(\widehat{R}\),

@@

We can interpret \(AC\)-values as a measure of non-stationarity. If there is dependence between subsequent \(\theta\)-values in imputation chains, \(AC\)-values are non-zero. Positive \(AC\)-values occur when \(\theta\)-values are recurring (i.e., high \(\theta\)-values are followed by high \(\theta\)-values, and low \(\theta\)-values are followed by low \(\theta\)-values). Recurrence within imputation chains may lead to trending. Negative \(AC\)-values occur when \(\theta\)-values of subsequent iterations are less similar, or diverge from one another. Divergence within imputation poses no threat to the convergence of the algorithm---it may even speed up convergence. \textbf{Complete stationarity is reached when \(AC=0\).} As non-convergence diagnostic, our interest is in positive \(AC\)-values.

@@

It is unlikely that iterative algorithms such as MICE will achieve the ideal values of \(\widehat{R}=1\) and \(AC=0\). Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state. The aim is therefore to reach approximate convergence \citep{gelm13}. Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm (\(\widehat{R} > 1\)), and there is little dependency between subsequent iterations of imputation chains (\(AC > 0\)). \textbf{Approximate convergence is violated when the non-convergence diagnostics exceed a certain threshold. {[}or{]} In practice, non-convergence is therefore diagnosed when non-convergence diagnostics exeed a certain threshold.} The conventional thresholds to diagnose non-mixing are \(\widehat{R} > 1.2\) \citep{gelm92} or \(\widehat{R} > 1.1\) \citep{gelm13}. \citet{veht19} proposed a much more stringent threshold of \(\widehat{R} > 1.01\). The magnitude of \(AC\)-values may be evaluated statistically, using a Wald test with \(AC=0\) as null hypothesis \citep{box15}. \(AC\)-values that are significantly higher than zero indicate non-stationarity. \textbf{{[}think about one-sided test{]}}

@@

Before we evaluate the performance of the non-convergence diagnostics \(\widehat{R}\) and \(AC\) quantitatively through simulation, we apply them to the example of pathological non-convergence by \citet{buur18} that we reproduced above. We want to be able to draw the same conclusions as we did from visually inspecting the two scenarios in Figure \ref{fig:non-conv}. The methods should at least indicate worse performance for the scenario with pathological non-convergence compared to the typically converged algorithm (i.e., higher \(\widehat{R}\)- and \(AC\)-values).

In Figure \ref{fig:diagnostics}A, the chain means from Figure \ref{fig:non-conv} are plotted again---now together. Panel B shows two versions of calculating \(AC\)s applied to the \(\theta\)s of panel A: the default calculation with \texttt{R} function \texttt{stats::acf()} \citep{R}, and manual calculation as the correlation between \(\theta\) in iteration \(t\) and \(\theta\) in iteration \(t+1\). Panel C shows the traditional computation of \(\widehat{R}\), and panel D shows \(\widehat{R}\) as computed by implementing \citet{veht19} 's recommendations. \textbf{{[}Add labels for the two types of Rhat to the figure!{]}}

>>>>

These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{zhu15, buur18}. There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{taka17}. Moreover, the behavoir of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}. Therefore, algorithmic convergence should be monitored carefully, \textbf{but this is not straightforward}. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence \citep{gelm13}. Since MCMC algorithms do not reach a unique point at which convergence is established, \textbf{methods} may only identify signs of \emph{non}-convergence \citep{hoff09}. Several non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.

In this paper, we study different methods for assessing non-convergence in iterative imputation algorithms. We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications. We then adress the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in \texttt{R} \citep{R}. For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R}. The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged. And additionally, to evaluate the behavior and performance of several diagnostic methods to identify non-convergence. With that, we formulate an informed advice on when approximate convergence, necessary for valid inferences, may be safely concluded. We translate the results of the study into guidelines for applied researchers. \textbf{ADD that these guidelines may aid to draw valid inference from incomplete data?{]}}

@@

Let \(y\) denote an \(n \times k\) matrix containing the data values on \(k\) variables for all \(n\) units in a sample. The data value of unit \(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\) (\(j = 1, 2, \dots, k\)) may be either observed or missing. The number of units \(i\) with at least one missing value, divided by the total number of units \(n\), is called the missingness proportion \(p_{\rm mis}\) in dataset \(y\). The collection of observed data values in \(y\) is denoted by \(y_{\rm obs}\); the missing part of \(y\) is referred to as \(y_{\rm mis}\). For each datapoint in \(y_{\rm mis}\), we sample \(m \times T\) plausible values, where \(m\) is the number of imputations (\(\ell = 1, 2, \dots, m\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm. The state space of the algorithm at a certain iteration \(t\) may be summarized by a scalar summary \(\theta\) (e.g., the average of the \textbf{imputed values}). The collection of \(\theta\)-values between \(t=1\) (summarizing the state space of the algorithm at initialization) and \(t=T\) (summarizing the state space for the imputed values) will be referred to as an `imputation chain'.

@@

There are two requirements for convergence of iterative algorithms: mixing and stationarity \citep{gelm13}. In iterative imputation algorithms, mixing implies that imputation chains intermingle nicely, and stationarity is characterized by the absence of trending across iterations. If one of the two requirements is not met, we speak of non-convergence. Without mixing, chains may be `stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values. The distribution of imputed values then differs across imputations. This may cause under-estimation of the variance between chains, which results in spurious, invalid inferences. Without stationarity, there is trending within imputation chains. Trending implies that further iterations would yield a systematically lower or higher set of imputations. Iterative imputation algorithms that have not (yet) reached stationarity, may thus yield biased estimates.

To illustrate what non-mixing and non-stationarity looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure \ref{fig:non-conv} displays two scenarios from the example. The graph on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. Each line portrays one imputation chain (i.e., the values of a scalar summary \(\theta\) across iterations). The \(\theta\) depicted here is the `chain mean' of variable \(j\), which is defined as the average of variable \(j\) in \(\dot y_{\rm imp, \ell}\).

In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence scenario, there is a lot of trending and some chains do not intermingle. Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two scenarios. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. It is obvious that non-convergence impacts the distribution of the imputed values in this example, \(\dot y_{\rm imp, \ell}\). This effect may translate into the distribution of the completed data \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}, which consequently affects the estimated quantities of scientific interest \(\hat{Q}_{\ell}\) and the pooled estimate \(\bar{Q}\). \(\bar{Q}\) is then a biased, invalid estimate of \(Q\), since systematically higher imputed values lead to bias, and under-estimation of the variance within imputation chains poses a threat to confidence-validity. Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms. \textbf{{[}ADD t and l to each theta??{]}}

@@

\caption{Typical convergence versus pathological non-convergence. Depicted are are chain means of variable $j$ in each imputation ($m=5$; $T=10$). Please note that any difference between the two scenarios (see e.g., the magnitude of the chain means on the y-axes) stems from the imputation algorithms---the two scenarios have equal incomplete data $y$.}\label{fig:non-conv}
\end{figure}

@@

Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection. After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots. In a traceplot, a scalar summary of the state space of the algorithm \(\theta\) is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}. In \texttt{mice}, the default scalar summaries to inspect are chain means and chain variances. Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends \citep{buur18}.

Identifying non-convergence by inspecting traceplots may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence \citep[\(\S\) 6.5.2]{buur18}. Moreover, the typical traceplots to inspect only \textbf{consider} univariate \(\theta\)s. Monitoring univariate summaries of the state space of the algorithm may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of the imputations. Ideally, we would monitor a multivariate \(\theta\).

A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the estimated quantity of scientific interest \(\hat{Q}\), which usually is multivariate in nature. This \(\theta\) is computed as the estimate in each imputation \(\ell\), \(\hat{Q}_\ell\), across iterations. Implementing this, however, might be somewhat too technical for empirical researchers. Besides, this scalar summary is not model-independent. That is, this \(\theta\) is not universal to all complete data problems, while one of the advantages of iterative imputation is that the missing data problem and scientific problem are solved independently. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest. On these grounds, such a \(\theta\) can be considered insufficient too.

@@

To monitor non-convergence in iterative imputation algorithms, we may summarize the state space of the algorithm with several \(\theta\)s. The downside to the current \(\theta\)s is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent \(\theta\) that summarizes the multivariate nature of the data.

We propose \(\lambda_{1}\) as such a scalar summary. We define \(\lambda_{1}\) as the first eigenvalue of the variance-covariance matrix in the \(m\) completed datasets. The eigenvalues of a variance-covariance matrix \(S\) summarize the total set of covariances in the data. Let \(\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}\) be the eigenvalues of \(S_\ell\) in each imputation \(\ell\). \(\lambda_{1,\ell}\) then summarizes the largest possible amount of covariance in the \(m\) completed datasets \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}. By definition, each \(\lambda_{1, \ell}\)-value is equal to the variance of the first component that would be obtained by performing principal component analysis on \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}. As scalar summary, \(\lambda_{1}\) has the appealing property that it is not dependent on the substantive model of interest, yet still summarizes the multivariate state space of the algorithm. \newline

\noindent In this study, we consider each of the four \(\theta\)s that were discussed above. Namely, two univariate scalar summaries (chain mean and chain variance), a model-dependent multivariate summary \(\hat{Q}\), and the novel model-independent multivariate summary \(\lambda_1\).

@@

There are many diagnostic tools to identify non-convergence in iterative (MCMC) algorithms \citep{broo98, elad06}. We consider only two of them that may be appropriate for imputation algorithms---one to monitor signs of non-mixing, and one for non-stationarity. As recommended by e.g.~\citet{cowl96} we will use the potential scale reduction factor \(\widehat{R}\) \citep[`Gelman-Rubin statistic';][]{gelm92} to evaluate mixing, and autocorrelation \citep[\(AC\);][]{scha97, gelm13} to diagnose trending. With a recently proposed adaptation, \(\widehat{R}\) might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated \citep{veht19}. Therefore, we will evaluate the appropriateness of both \(\widehat{R}\) versions, in addition to \(AC\). Other methods are outside the scope of this study, because they e.g.~assume that values within chains represent independent samples, whereas MI only uses the final iteration to impute.

@@

In 2019, Vehtari et al.~proposed an updated version of the potential scale reduction factor \(\widehat{R}\), originally coined in 1992. The adapted version would be better suited to detect non-mixing in the tails of distributions, and even identify non-stationarity. This version uses three transformations on the scalar summary \(\theta\), before computing \(\widehat{R}\)-values. Namely, rank-normalization, folding, and localization. The definition of the diagnostic itself is equal to the original \(\widehat{R}\). We follow Vehtari et al.'s formulation (2019, p.~5). Let \(m\) be the total number of chains, \(T\) the number of iterations per chain (where \(T\geq2\)), and \(\theta\) the scalar summary of interest. For each chain (\(\ell = 1, 2, \dots, m\)), we estimate the variance of \(\theta\), and average these to obtain within-chain variance \(W\).

@@

\noindent We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\)s per chain. \textbf{Note that \(B\) refers to the variance in \(\theta\), instead of the typical notation in MI where \(B\) denotes the variance between estimated quantities of scientific interest \(\hat{Q}_{\ell}\).}

@@

Autocorrelation is defined as the correlation between two subsequent \(\theta\)-values within the same chain \citep[p.~147]{lync07}. In this study, we only consider \(AC\) at lag 1, i.e., the correlation between the \(t^{th}\) and \((t+1)^{th}\) iteration of the same chain. Following the same notation as for \(\widehat{R}\),

@@

We can interpret \(AC\)-values as a measure of non-stationarity. If there is dependence between subsequent \(\theta\)-values in imputation chains, \(AC\)-values are non-zero. Positive \(AC\)-values occur when \(\theta\)-values are recurring (i.e., high \(\theta\)-values are followed by high \(\theta\)-values, and low \(\theta\)-values are followed by low \(\theta\)-values). Recurrence within imputation chains may lead to trending. Negative \(AC\)-values occur when \(\theta\)-values of subsequent iterations are less similar, or diverge from one another. Divergence within imputation poses no threat to the convergence of the algorithm---it may even speed up convergence. Complete stationarity is reached when \(AC=0\). As non-convergence diagnostic, our interest is in positive \(AC\)-values.

@@

It is unlikely that iterative algorithms such as MICE will achieve the ideal values of \(\widehat{R}=1\) and \(AC=0\). Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state. The aim, therefore, is to reach approximate convergence \citep{gelm13}. Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm (\(\widehat{R} > 1\)), and there is little dependency between subsequent iterations of imputation chains (\(AC > 0\)). In practice, we diagnose non-convergence when approximate convergence is violated, i.e., when \(\widehat{R}\) and \(AC\) exceed a certain threshold. The conventional thresholds to diagnose non-mixing are \(\widehat{R} > 1.2\) \citep{gelm92} or \(\widehat{R} > 1.1\) \citep{gelm13}. \citet{veht19} proposed a much more stringent threshold of \(\widehat{R} > 1.01\). The magnitude of \(AC\)-values may be evaluated statistically, using a Wald test with \(AC=0\) as null hypothesis \citep{box15}. \(AC\)-values that are significantly higher than zero indicate non-stationarity.

@@

Before we evaluate the performance of the non-convergence diagnostics \(\widehat{R}\) and \(AC\) quantitatively through simulation, we first assess their appropriateness qualitatively. We do this by applying them to the example of pathological non-convergence that we reproduced form \citet{buur18}. Ideally, the diagnostics are as informative as, or better than visual inspection of the traceplots. The methods should at least indicate worse performance (higher \(\widehat{R}\)- and \(AC\)-values) for the scenario with pathological non-convergence, compared to the typically converged algorithm.

Each panel in Figure \ref{fig:diagnostics} depicts one method to identify non-convergence, applied to the two scenarios from Figure \ref{fig:non-conv}. Panel A consists of the two traceplots that may be evaluated through visual inspection. Panel B shows two versions of the \(AC\): the default calculation with \texttt{R} function \texttt{stats::acf()} \citep{R}, and manual calculation as the correlation between \(\theta\) in iteration \(t\) and \(\theta\) in iteration \(t+1\). Panel C displays the traditional computation of \(\widehat{R}\) conform \citet{gelm92}, and in panel D we see \(\widehat{R}\) as computed by implementing \citet{veht19} 's recommendations. \textbf{{[}Add labels for the two types of Rhat to the figure!{]}}
